\chapter{Counting}\label{chap:counting}
In this chapter, we will explore the concept of counting in the context of formal languages and automata theory as well as explain the implementation for fixed and bounded repetition in FAdo.

\section{Counting in Formal Languages}
In formal language theory, counting refers to the ability of a language or an automaton to enforce numeric constraints over the number of symbols or patterns within strings. Specifically, it deals with the ability to recognize whether certain elements occur a specified number of times—or in a specific numerical relationship to others.
Current tools are already able to do this, including some non-backtracking matchers.

\section{Derivatives of operations in Extended Regular Expressions} %TODO: CHANGE THIS
Given a word $w = xyz$ and a regular expression $\alpha$, $w \in L(\alpha)$ if $\varepsilon(d_w(\alpha)) = \varepsilon$.
In order to match with fixed and bounded repetition, one must first solve the derivatives for those operations.

\begin{thm}
	Given $e \in RegExp(\Sigma)$, the following holds:
	\begin{align*}
		\varepsilon(e) + e = e
	\end{align*}
\end{thm}
\begin{proof}
	Let $e \in RegExp(\Sigma)$. The proof is by iterating through both possible scenarios where $\varepsilon \in L(e)$ and $\varepsilon \notin L(e)$.
	For $\varepsilon \in L(e)$, we have that $\varepsilon(e) = \varepsilon$. And so,
	\begin{align*}
		& \varepsilon(e) + e = e \\
		& \Rightarrow \varepsilon + e = e \\
		& \Rightarrow e = e
	\end{align*}
	holds. On the other hand, for $\varepsilon \notin L(e)$, we have that $\varepsilon(e) = \emptyset$. Therefore,
	\begin{align*}
		& \varepsilon(e) + e = e \\
		& \Rightarrow \emptyset + e = e \\
		& \Rightarrow e = e
	\end{align*}
	also holds.
\end{proof}

\subsection{Fixed Repetition}
For a fixed repetition, we need $r$ to occur $n$ times, resulting in the notion $r^n$.
% PROOF
%\begin{thm}
%	Given $e \in RegExp(\Sigma)$ and $n \in \Npos$, we have that
%	\begin{align*}
%		d_a(e^n) = d_a(e)e^{n-1}
%	\end{align*}
%	is true.
%\end{thm}
%\begin{proof}
%	Let $e \in RegExp(\Sigma)$ and $n \in \Npos$. The following holds:
%	\begin{align*}
%		d_a(e) = d_a(e)e^0 = d_a(e)
%	\end{align*}
%	We can also suppose that
%	\begin{align*}
%		d_a(e^{n-1}) = d_a(e)e^{n-2}
%	\end{align*}
%	And so, we can use the results above to do the following:
%	\begin{align*}
%		d_a(e^n) &= d_a(ee^{n-1}) \\
%		&= d_a(e)e^{n-1} + \varepsilon(e)d_a(e^{n-1}) \\
%		&= d_a(e)e^{n-1} + \varepsilon(e)d_a(e)e^{n-2} \\
%		&= (d_a(e)e + \varepsilon(e)d_a(e))e^{n-2} \\
%		&= (d_a(e)(e+\varepsilon(e)))e^{n-2} \\
%		&= d_a(e)e^{n-1}
%	\end{align*}
%\end{proof}
\begin{thm}
	\label{thm:fixed_repetition_derivative}
	Given $e \in RegExp(\Sigma)$ and $n \in \Npos$, we have:
	\begin{align*}
		d_a(e^n) = d_a(e)\,e^{n-1}.
	\end{align*}
\end{thm}

\begin{proof}
	We proceed by induction on $n \in \Npos$. For the base case \( n = 1 \),
	\begin{align*}
		d_a(e^1) = d_a(e) = d_a(e)e^0 = d_a(e)\varepsilon = d_a(e).
	\end{align*}
	Since the identity holds for \(n = 1\), lets assume that the identity holds for some \(n-1 \geq 1\), i.e.,
	\begin{align*}
		d_a(e^{n-1}) = d_a(e)e^{n-2}.
	\end{align*}
	We want to show it holds for \(n\). Using the product rule:
	\begin{align*}
		d_a(e^n) &= d_a(ee^{n-1}) \\
		&= d_a(e) e^{n-1} + \varepsilon(e) d_a(e^{n-1}) \\
		&= d_a(e) e^{n-1} + \varepsilon(e) d_a(e) e^{n-2} \\
		&= d_a(e) \left(e^{n-1} + \varepsilon(e) e^{n-2}\right) \\
		&= d_a(e) \left(ee^{n-2} + \varepsilon(e) e^{n-2}\right) \\
		&= d_a(e) (e + \varepsilon(e))e^{n-2} \\
		&= d_a(e) e^{n-1}.
	\end{align*}
	This completes the induction.
\end{proof}

As an example, let $r = ab$ such that $r^2 = abab$. Replacing with the theorem above:
\begin{align*}
	d_a(r^2) &= d_a(r \cdot r) \\
	&= d_a(abab)  \\
	&= d_a(a)bab + d_a(bab) \\
	&= \varepsilon bab + \emptyset \\
	&= bab
\end{align*}

% ADD e^n results for n=0, n=1 and n>1

\subsection{Bounded Repetition}
Supplementing the fixed repetition notion, we now add a maximum bound so that $r$ can occur at least $n$ times and at maximum $m-1$ times: $r^{n,m}$.

\begin{thm}
	Given $e \in RegExp(\Sigma)$, $n \in \Npos$, $m > n$, and $a \in \Sigma$, we have:
	\begin{align*}
		d_a(e^{n,m}) =
		\begin{cases}
			d_a(e) e^{n-1,m-1}, & \text{if } n > 1, \\
			d_a(e) + d_a(e) e^{1,m-1}, & \text{if } n = 1.
		\end{cases}
	\end{align*}
\end{thm}

\begin{proof}
	Let $e \in RegExp(\Sigma)$ and let $n, m \in \Npos$ with $m > n$. By definition of bounded iteration:
	\begin{align*}
		e^{n,m} &= \sum_{i=n}^{m} e^i,
	\end{align*}
	and by the linearity of derivatives over sums:
	\begin{align*}
		d_a(e^{n,m}) &= \sum_{i=n}^{m} d_a(e^i).
	\end{align*}
	Using the known identity $d_a(e^i) = d_a(e) e^{i-1}$, we have:
	\begin{align*}
		d_a(e^{n,m}) &= \sum_{i=n}^{m} d_a(e) e^{i-1} \\
		&= d_a(e) \sum_{i=n}^{m} e^{i-1}.
	\end{align*}
	Letting $j = i - 1$, this becomes:
	\begin{align*}
		d_a(e^{n,m}) &= d_a(e) \sum_{j=n-1}^{m-1} e^j = d_a(e) e^{n-1,m-1}.
	\end{align*}
	Considering both cases, for $n > 1$, the expression is already in its final form:
	\begin{align*}
		d_a(e^{n,m}) = d_a(e) e^{n-1,m-1}.
	\end{align*}
	And for $n = 1$, we have:
	\begin{align*}
		d_a(e^{1,m}) &= d_a(e) e^{0,m-1} \\
		&= d_a(e) + d_a(e) e^{1,m-1}.
	\end{align*}
	This completes the proof.
\end{proof}
Furthermore, we can include the infinite upper bound: $r^{n,\infty}$.
\begin{align*}
	r^{n,\infty} = r^nr^\star
\end{align*}
We can simplify it even more, depending on the value of $n$. Assuming $n=0$:
\begin{align*}
	r^{0,\infty} = r^0r^\star = \varepsilon r^\star =  r^\star
\end{align*}
Lastly, for $n=1$:
\begin{align*}
	r^{1,\infty} = r^1r^\star = rr^\star =  r^+
\end{align*}

\begin{thm}
	Given $e \in RegExp(\Sigma)$, $n \in \N$ and $a \in \Sigma$, we have:
	\begin{align*}
		d_a(e^{n,\infty}) =
		\begin{cases}
			d_a(e)e^{n-1}e^\star, & \text{if } n>0, \\
			d_a(e)e^*, & \text{otherwise}.
		\end{cases}
	\end{align*}
\end{thm}
\begin{proof}
	Let $e \in RegExp(\Sigma)$, $n \in \N$, and $a \in \Sigma$. We have,
	\begin{align*}
		d_a(e^{n,\infty}) &= d_a(e^ne^\star) \\
		&= d_a(e^n)e^\star. %\\
%		\intertext{using \ref{thm:fixed_repetition_derivative}:}
%		&= d_a(e)e^{n-1}e^\star
	\end{align*}
	Now, if $n>0$, we can use \ref{thm:fixed_repetition_derivative}:
	\begin{align*}
		d_a(e^n)e^\star &= d_a(e)e^{n-1}e^\star.
	\end{align*}
	Otherwise, if $n=0$, we can just replace accordingly:
	\begin{align*}
		d_a(e^ne^\star) = d_a(e^0e^\star) = d_a(e^*) = d_a(e)e^*.
	\end{align*}
	This concludes the proof.
\end{proof}
For instance, given $r = ab$ such that $r^2 = abab$ and $r^3 = ababab$, we have:
\begin{align*}
	d_a(r^{[2,4[}) &= d_a(abab) + d_a(ababab) \\
	&= d_a(abab) + d_a(ababab) \\
	&= d_a(a)bab + d_a(bab) + d_a(a)babab + d_a(babab) \\
	&= \varepsilon bab + \emptyset + babab + \emptyset \\
	&= bab + babab.
\end{align*}
Extending to the infinity case, $r^{[2, \inf[}$:
\begin{align*}
	d_a(r^{[2,\inf[}) &= d_a(abab) + d_a((ab)^*) \\
	&= d_a(a)bab + d_a(bab) + d_a(ab)\cdot(ab)^* \\
	&= \varepsilon bab + \emptyset + (d_a(a)b + d_a(b)) \cdot (ab)^* \\
	&= bab + (\varepsilon b + \emptyset) \cdot (ab)^* \\
	&= bab + b(ab)^*.
\end{align*}

%For instance, if $m = 2$, we have that $r^{[2, \inf[}$

%The regular expression $r^{[2, \inf[}$ will match any $(ab)^k$ with $k \geq 2$.

%Given $w = abababab$ and $m = 2$,  will match 

%The string $abababab$ can be matched by $r^{[2, \inf[}$

%\begin{thm}
%	Given $e \in RegExp(\Sigma)$, $n \in \Npos$ and $m>n$, we have that
%	\begin{align*}
%	%	d_a(e^{n,m}) = d_a(e) e^{n-1,m-1}
%	d_a(e^{n,m}) = \begin{cases}
%		d_a(e)e^{n-1,m-1} & n>1 \\
%		d_a(e)+(d_a(e))e^{1,m-1} & n=1
%	\end{cases}
%	\end{align*}
%	is true.
%\end{thm}
%\begin{proof}
%	Let $e \in RegExp(\Sigma)$. We first need to simplify the main expression:
%	\begin{align*}
%		d_a(e^{n,m}) &= \sum_{i=n}^{m} d_a(e^i) \\
%		&= \sum_{i=n}^{m} (d_a(e))e^{i-1} \\
%		&= d_a(e) \sum_{j=n-1}^{m-1} e^j \\
%		&= d_a(e) e^{n-1,m-1}
%	\end{align*}
%	Thus, for $n=1$, we have that
%	\begin{align*}
%		d_a(e^{n,m}) &= d_a(e)e^{0,m-1} \\
%		&= d_a(e) + (d_a(e))e^{1,m-1}
%	\end{align*}
%	And lastly, for $n>=1$ we have that
%	\begin{align*}
%		d_a(e^{n,m}) &= (d_a(e))e^{n-1,m-1}
%	\end{align*}
%\end{proof}

%For instance, if $m = 2$, we have that $r^{[2, \inf[}$

%The regular expression $r^{[2, \inf[}$ will match any $(ab)^k$ with $k \geq 2$.

%Given $w = abababab$ and $m = 2$,  will match 

%The string $abababab$ can be matched by $r^{[2, \inf[}$


\section{Implementation in FAdo}
In order to implement these new syntactic constructs of extended regular expressions in FAdo, we added exact-power and counted-repetition nodes to the regular expression abstract syntax tree and define their derivatives, partial derivatives and linear form cases.

\subsection{CPower}
\textit{CPower} is the class responsible for the construction of regular expressions using the fixed repetition operator.

\begin{lstlisting}[language=Python]
class CPower(Unary):
	def __init__(self, arg, n, sigma=None):
	self.arg = arg
	self.n = n
	self.Sigma = sigma
	self._ewp = False if self.n > 0 else True
	
	...
\end{lstlisting}

As shown above, the class inherits from the \textit{Unary}, which only defines the \textit{Unary.Sigma} and \textit{Unary.arg} properties, letting the class hold a alphabet set (representing $\Sigma$) and the symbol used to construct this operator.

In order to integrate the class fully with \textit{FAdo}, some more methods had to be implemented.

\begin{lstlisting}[language=Python]
	def linearForm(self):
		arg_lf = self.arg.linearForm()
		lf = dict()
		for head in arg_lf:
			lf[head] = set()
			for tail in arg_lf[head]:
				lf[head].add(self.derivative(head))
		return lf

	def partialDerivatives(self, sigma):
		return self.arg.partialDerivatives(sigma)
	
	def derivative(self, sigma):
		if str(sigma) in str(self.Sigma):
			if self.n == 0:
				return CEmptySet(sigma)
			elif self.n == 1:
				return self.arg.derivative(sigma)
			else:
				return CConcat(self.arg.derivative(sigma), CPower(self.arg, self.n-1, self.Sigma))
		else:
			return CEmptySet(sigma)
\end{lstlisting}

\subsection{CCount}
\textit{CCount} is the class responsible for the construction of regular expressions using the bounded repetition operator.


\begin{lstlisting}[language=Python]
	class CCount(Unary):
		def __init__(self, arg, min, max = None, sigma = None):
			self.arg = arg
			self.min = int(min)
			self.max = "inf" if max == -1 or max == "inf" else int(max)-1
			self.Sigma = sigma
	
			if self.min==0:
				self._ewp = True
\end{lstlisting}

As shown above, and much like \textit{CPower}, the class also inherits from \textit{Unary}. The constructor adjusts the bounds accordingly and sets the empty word property based on whether the minimum repetition is zero.

To fully integrate the class with \textit{FAdo}, several methods are implemented, including string representation, partial derivatives, linear form computation, and marking functions:

\begin{lstlisting}[language=Python]
	def derivative(self, sigma):
		if str(sigma) in str(self.Sigma):
			if self.max == "inf" or self.max == None:
				if self.min == 0:
					return CConcat(self.arg.derivative(sigma), CStar(self.arg, self.Sigma))
				else:
					return CConcat(self.arg.derivative(sigma), CCount(self.arg, self.min-1, self.max, sigma), self.Sigma)
			else:
				if self.max > 1:
					if self.min == 0:
						return CConcat(self.arg.derivative(sigma), CCount(self.arg, self.min, int(self.max), self.Sigma))
					else:
						return CConcat(self.arg.derivative(sigma), CCount(self.arg, self.min-1, int(self.max), self.Sigma))
				elif self.max == 1:
					return self.arg.derivative(sigma)
				else:
					CEpsilon(sigma)
		else:
			return CEmptySet(sigma)
		
	def partialDerivatives(self, sigma):
		arg_pdset = self.arg.partialDerivatives(sigma)
		pds = set()
		for pd in arg_pdset:
			if pd.emptysetP():
				pds.add(CEmptySet(self.Sigma))
			elif pd.epsilonP():
				pds.add(self)
			else:
				pds.add(CConcat(pd, self, self.Sigma))
		return pds
\end{lstlisting}

The \texttt{linearForm} method is exactly the same as \texttt{CPower}'s.

%\begin{lstlisting}[language=Python]
%	def linearForm(self):
%		arg_lf = self.arg.linearForm()
%		lf = dict()
%		for head in arg_lf:
%			lf[head] = set()
%			for tail in arg_lf[head]:
%				lf[head].add(self.derivative(head))
%		return lf
%\end{lstlisting}

\subsection{Grammar}
For ease of use and protoyping, we utilized \textit{FAdo}'s ability to parse a regular expression using a common Python \textit{string}. To do this, \textit{FAdo} utilizes a library called Lark. Lark is a modern, fast and flexible parsing toolkit for Python that supports both LALR(1) and Earley parsers, enabling the parsing of all context-free grammars with features like EBNF, AST generation, and grammar composition \cite{lark_parser}.
The new grammar can be seen in \ref{appendix:regexp_lark} and the most notable changes are the following:
\begin{lstlisting}
	| rep _CARET _LEFBR digit _RIGBR -> pow_min
	| rep _CARET _LEFBR digit _COMMA digit _LEFBR -> pow_minmax
	| rep _CARET _LEFBR digit _COMMA _INFTY _LEFBR -> pow_inf
\end{lstlisting}
These rules (along with some more utilitary definitions such as the \texttt{\_LEFBR} and \texttt{\_RIGBR}, which are the symbols '\texttt{\string[}' and '\texttt{\string]}') will route the following cases (code in Appendix \ref{appendix:regexp_builders}):
\begin{itemize}
	\item \textbf{pow\_min}: For the cases $r^{n}$ and $r^{n,\infty}$, where $r \in RegExp(\Sigma)$ and $n \in \Npos$
	\item \textbf{pow\_minmax}: For the case $r^{n,m}$, where $r \in RegExp(\Sigma)$ and $n \in \Npos, m>n$
	\item \textbf{pow\_inf}: Calls \textit{pow\_min} with a special argument that selects the infinity case
\end{itemize}

\section{Single Matching}
When referring to matching in the context of regular expressions, one typically means simply determining whether a given word is a member of the language defined by a specific regular expression.
We can define this problem formally as:
\begin{defn}
	\label{defn:single_matching}
	Given $e \in RegExp(\Sigma)$ and a word $w \in \Sigma^\star$, we say that $w$ matches $e$ if $w \in L(e)$.
\end{defn}

To look for matches using \textit{FAdo}, we first need to choose a regular expression. For the sake of this example, let $r = ab + ba$. Informally, this regular expression will match any $ab$ or $ba$ character sequences in an input text. To do this, we first need to import the grammar using Lark and parse the regular expression to get a parsed syntax tree. With this tree, we can use \textit{FAdo}'s \texttt{BuildRegexp} class and invoke its \texttt{transform} method, which will iteratively transform each of the trees nodes into abstract \textit{FAdo} class objects:
\begin{lstlisting}[language=Python]
	r = "ab+ba"
	regGrammar = lark.Lark.open("lang/regexp_test.lark", start="rege", parser="lalr")
	tree = regGrammar.parse(r)
	reg : RegExp = BuildRegexp(context={"sigma": None}).transform(tree)
	reg.setSigma(reg.setOfSymbols())
\end{lstlisting}
We now have a regular expression object. We can use it to build a DFA using Brzozowski's derivatives construction:
\begin{lstlisting}[language=Python]
	dfa_z = reg.dfaBrzozowski()
\end{lstlisting}
Finally, having a DFA object, we can then use the \texttt{evalWordP} function to verify if $w \in L(r)$ for a given $w \in \Sigma^\star$, which can be, for this example, $w = ab$:
\begin{lstlisting}[language=Python,label={code:eval_word_dfa}]
	w = "ab"
	if dfa_z.evalWordP(w):
		print("w belongs to the language of r")
	else:
		print("w does not belong to the language of r")
\end{lstlisting}
With the result being:
\begin{lstlisting}[language=Python]
	w belongs to the language of r
\end{lstlisting}
We can do the exact same with the new syntax. Let $r = (ab)^{1,3}$. First, parse the regular expression and display it:
\begin{lstlisting}[language=Python]
	r = "(ab)^[1,3["
	tree = regGrammar.parse(r)
	reg : RegExp = BuildRegexp(context={"sigma": None}).transform(tree)
	reg.setSigma(reg.setOfSymbols())
	dfa_z = reg.dfaBrzozowski()
	dfa_z.display()
\end{lstlisting}

\begin{figure}[H]
	\centering
	\begin{tikzpicture}[shorten >=1pt,node distance=2cm,on grid, auto,initial text=]
		
		% --- States ---
		\node[state, initial] (q0) {0};
		\node[state]          (q1) [above right=of q0] {1};
		\node[state]          (q2) [below right=of q0] {2};
		\node[state, accepting]         (q3) [right=22mm of q1] {3};
		\node[state]          (q4) [right=22mm of q2] {4};
		\node[state]          (q5) [above right=of q3] {5};
		\node[state]          (q6) [right=20mm of q3] {6};
		\node[state, accepting]         (q7) [right=22mm of q6] {7};
		\node[state]          (q8) [right=22mm of q7] {8};
		
		% --- Transitions ---
		\path[->]
		(q0) edge node {$b$} (q1)
		edge[swap] node {$a$} (q2)
		
		(q1) edge[loop above] node {$b, a$} ()
		
		(q2) edge node {$b$} (q3)
		edge[swap] node {$a$} (q4)
		
		(q4) edge[loop below] node {$b, a$} ()
		
		(q3) edge node {$b$} (q5)
		edge[swap] node {$a$} (q6)
		
		(q5) edge[loop above] node {$b, a$} ()
		
		(q6) edge node {$b$} (q7)
		edge[bend right=40, swap] node {$a$} (q8)
		
		(q7) edge node {$b, a$} (q8)
		
		(q8) edge[loop right] node {$b, a$} ();
	\end{tikzpicture}
	\caption{Diagram of the DFA built using Brzozowski's method for the regular expression $r = (ab)^{1,3}$.}
	\label{fig:dfa_brzozowski_counting}
\end{figure}
Then, try to match $w = ab$ using the same code from \ref{code:eval_word_dfa}, which yields the following:
\begin{lstlisting}[language=Python]
	w belongs to the language of r
\end{lstlisting}
Trying again with a text that does not match, $w = ababab$, yields:
\begin{lstlisting}[language=Python]
	w does not belong to the language of r
\end{lstlisting}
%Formally, .